Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-12-11 manuscript/manuscript.tex
Minor improvements in methodology

\begin{equation}
    g_{\alpha\beta}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') I_{\alpha\beta} \, \kappa \, dr' d\phi' d\lambda' ,
    \label{eq:tesseroid-tensor}
\end{equation}

\noindent in which

\begin{equation}
    I_{\alpha\beta} =
    \left(
        \frac{3\Delta_{\alpha} \Delta_{\beta}}{\ell^5} -
        \frac{\delta_{\alpha\beta}}{\ell^3}
    \right) ,
    \label{eq:tesseroid-tensor-kernel}
\end{equation}

\noindent $\alpha, \beta \in \{x, y, z\}$, $\delta_{\alpha\beta}$ is Kronecker's delta,
$G = 6.674\times10^{-11}\, \text{m$^3$kg$^{-1}$s$^{-1}$}$ is the gravitational constant
and

@@

Applying a $N$th order GLQ, we can approximate each integral in equations
\ref{eq:tesseroid-pot}, \ref{eq:tesseroid-grav} and \ref{eq:tesseroid-tensor} by a
weighted sum of the integration kernel evaluated on the roots of an $N$th order Legendre
polynomial \citep[p.~390]{Hildebrand1987}.
Unlike the homogeneous density case, the radial density function $\rho(r)$ must also be
included in the integration and evaluated on the Legendre polynomial roots (i.e.,
quadrature nodes).

@@

\noindent $f(r', \phi', \lambda')$ is the non-density dependent portion of the
integration kernels, $(r_i, \phi_j, \lambda_k)$ are the quadrature nodes, and
$W_i^r$, $W_j^\phi$, $W_k^\lambda$ are the quadrature weights.
The GLQ nodes and weights can be calculated as \citet[p.~391]{Hildebrand1987} (see
\citet{Uieda2016} for nodes and weights definitions on similar notation as this paper).

@@

\citet{Ku1977} noticed that the GLQ integration
becomes less accurate when the computation point is closer to the
mass element.
One way to prevent this from happening would be to increase the GLQ order.
Doing so would uniformly increase the number of point masses inside the
tesseroid volume.
However, an increase in the point mass concentration is only required close to the
computation point \citep{Uieda2016}.
Alternatively, \citet{Li2011} proposed an adaptive
discretization algorithm which keeps the GLQ order fixed and divides the
tesseroid based on a ratio between the distance to the computation
point and its dimensions.
This algorithm produces a more efficient computation because an increased concentration
of point masses is produced where it is needed more.
\citet{Uieda2016} developed a modified version of this algorithm that speeds the
computation of the distances.
Both algorithms (the original one from \citet{Li2011} and the modified
version of \citet{Uieda2016}) performs tesseroid subdivisions in the latitudinal,
longitudinal and radial directions, thus we can define them as three dimensional
adaptive discretization algorithms.
On the other hand, \citet{Lin2018} proposed a two dimensional discretization algorithm
that subdivides the tesseroid only on the latitudinal and longitudinal directions.
Removing a dimension from the splitting makes the computation more efficient, while
keeping an acceptable accuracy on the gravitational fields \citep{Lin2018}.

Here we will use a two dimensional version of the adaptive discretization algorithm
developed by \citet{Uieda2016}.
What follows is a summary of the algorithm and the reader is referred to
\citet{Uieda2016} for a detailed description.

Lets consider a computation point located at $(r, \phi, \lambda)$ and an arbitrary
tesseroid with its geometric centre at $(r_t, \phi_t, \lambda_t)$.
We can define its longitudinal ($L_\lambda$) and latitudinal ($L_\phi$) dimensions as
follows:

@@

\textit{Final step}:
By the end of the algorithm, the running total will be the gravitational effect of the
tesseroid.

Notice that the distance-size ratio $D$ determines how many
times the tesseroids will be divided.
Therefore, it effectively regulates both the accuracy of the algorithm and its
computation time.
An optimal value for $D$ cannot be directly calculated from the desired accuracy level.
Instead, it is empirically determined by comparing the numerical results with the
analytical solution for a spherical shell.
\citet{Uieda2016} used a shell with homogeneous density to determine optimal values of
$D$.
Here, we will repeat the numerical experiment using analytical expressions for shells
with density varying according to exponential and linear functions of $r$.
This numerical experiment will also show that the two dimensional adaptive
discretization algorithm is as accurate as the original three dimensional version,
though it reduces the number of tesseroid subdivisions and thus the computation time.

>>>>

\noindent in which $\alpha, \beta \in \{x, y, z\}$, $\delta_{\alpha\beta}$ is
Kronecker's delta, $G = 6.674\times10^{-11}\, \text{m$^3$kg$^{-1}$s$^{-1}$}$ is the
gravitational constant and

@@

Applying a $N$th order GLQ, we can approximate each integral in equations
\ref{eq:tesseroid-pot} and \ref{eq:tesseroid-grav} by a weighted sum of the integration
kernel evaluated on the roots of an $N$th order Legendre polynomial
\citep[p.~390]{Hildebrand1987}.
Unlike the homogeneous density case, the radial density function $\rho(r)$ must also be
included in the integration and evaluated on the Legendre polynomial roots (i.e.,
quadrature nodes).

@@

\noindent $f(r', \phi', \lambda')$ is the non-density dependent portion of the
integration kernels, $(r_i, \phi_j, \lambda_k)$ are the quadrature nodes, and
$W_i^r$, $W_j^\phi$, $W_k^\lambda$ are the quadrature weights.
The GLQ nodes and weights can be calculated as \citet[p.~391]{Hildebrand1987} (see
\citet{Uieda2016} for nodes and weights definitions on similar notation used on previous
equations).

It's worth noting that a GLQ is equivalent to approximate the tesseroid by $N^r \times
N^\phi \times N^\lambda$ point masses located on the quadrature nodes
\citep{Ku1977, Li2011, Uieda2016}.

@@

\citet{Ku1977} noticed that the GLQ integration
becomes less accurate when the computation point is closer to the
mass element.
One way to prevent this from happening would be to increase the GLQ order.
Doing so would uniformly increase the number of point masses inside the
tesseroid volume.
However, an increase in the point mass concentration is only required close to the
computation point \citep{Uieda2016}.
Alternatively, \citet{Li2011} proposed an adaptive
discretization algorithm which keeps the GLQ order fixed and divides the
tesseroid based on a ratio between the distance to the computation
point and its dimensions.
This algorithm produces a more efficient computation because an increased concentration
of point masses is produced where it is needed more.
\citet{Uieda2016} developed a modified version of this algorithm that speeds the
computation of the distances.
Both algorithms (the original one from \citet{Li2011} and the modified
version of \citet{Uieda2016}) perform tesseroid subdivisions in the latitudinal,
longitudinal and radial directions, thus we can define them as three dimensional
adaptive discretization algorithms.
On the other hand, \citet{Lin2018} proposed a two dimensional discretization algorithm
that subdivides the tesseroid only on the latitudinal and longitudinal directions.
Removing a dimension from the splitting makes the computation more efficient, while
keeping an acceptable accuracy on the gravitational fields \citep{Lin2018}.

Here we will use a two dimensional version of the adaptive discretization algorithm
developed by \citet{Uieda2016}: the distances from the tesseroid and the computation
point are computed as \citet{Uieda2016} do, but the tesseroid will only be split on the
longitudinal and latitudinal directions.
What follows is a summary of the algorithm and the reader is referred to
\citet{Uieda2016} for a detailed description.

Lets consider a computation point located at $(r, \phi, \lambda)$ and an arbitrary
tesseroid defined by the boundaries $r_1$, $r_2$, $\phi_1$, $\phi_2$, $\lambda_1$ and
$\lambda_2$, which geometric centre is located at $(r_t, \phi_t, \lambda_t)$, where

\begin{equation}
    r_t = \frac{r_2 + r_1}{2}, \quad
    \phi_t = \frac{\phi_2 + \phi_1}{2}, \quad
    \lambda_t = \frac{\lambda_2 + \lambda_1}{2}.
\end{equation}

We can define the longitudinal ($L_\lambda$) and latitudinal ($L_\phi$) dimensions of
the tesseroid as follows:

@@

\textit{Final step}:
By the end of the algorithm, a second-order GLQ will be applied on each small tesseroid
to compute the gravitational field they generate, and the sum of these results will be
the gravitational effect of the original tesseroid.

Notice that the distance-size ratio $D$ determines how many
times the tesseroids will be divided.
Therefore, it effectively regulates both the accuracy of the algorithm and its
computation time.
An optimal value for $D$ cannot be directly calculated from the desired accuracy level.
Instead, it is empirically determined by comparing the numerical results with the
analytical solution for a spherical shell.
\citet{Uieda2016} used a shell with homogeneous density to determine optimal values of
$D$.
Here, we will repeat the numerical experiment using analytical expressions for shells
with density varying according to certain functions of $r$.
This numerical experiment will also show that the two dimensional adaptive
discretization algorithm can achieve the same accuracy as the original three dimensional
version, though it reduces the number of tesseroid subdivisions and thus the computation
time.
